Generating gaits for physical quadruped robots: evolved neural networks vs. local parameterized search

### 2011-04-14 paper_ecal/conclusion.tex
more more forging

We have presented an array of approaches to optimizing a fast quadrupedal
gait.  We implemented and tested six learning
strategies for parameterized gaits and compared them to gaits produced by neural networks
evolved with the HyperNEAT generative encoding.

@@

HyperNEAT produced higher-performing gaits than all of the parameterized
methods. Its best-performing gait traveled 45.7 body lengths per minute, which is nearly 9 times the speed of the hand-coded gait.  This could be because HyperNEAT tends to generate coordinated gaits~\cite{clune2011performance, clune2009evolving}, allowing it to
take advantage of the symmetries of the problem, but can also explore a much larger space of possibilities than the
more restrictive 5-dimensional parameterized space.  HyperNEAT gaits tended to produce more complex sequences of motor commands, with different frequencies and degrees of coordination, whereas the
parameterized gaits were restricted to scaling single-frequency sine waves and could only produce certain types of motor regularities.

>>>>

We have presented an array of approaches for optimizing a quadrupedal
gaits for speed.  We implemented and tested six learning
strategies for parameterized gaits and compared them to gaits produced by neural networks
evolved with the HyperNEAT generative encoding.

@@

HyperNEAT produced higher-performing gaits than all of the parameterized
methods. Its best-performing gait traveled 45.7 body lengths per minute, which is nearly 9 times the speed of the hand-coded gait.  This could be because HyperNEAT tends to generate coordinated gaits~\citep{clune2011performance, clune2009evolving}, allowing it to
take advantage of the symmetries of the problem. HyperNEAT can also explore a much larger space of possibilities than the
more restrictive 5-dimensional parameterized space.  HyperNEAT gaits tended to produce more complex sequences of motor commands, with different frequencies and degrees of coordination, whereas the
parameterized gaits were restricted to scaling single-frequency sine waves and could only produce certain types of motor regularities.

### 2011-04-14 paper_ecal/futureWork.tex
more more forging

Because all trials were done in hardware, it was difficult to gather
the many trials that would be necessary to properly rank the methods statistically.
One direction for future work could be to obtain many more trials. However, a more effective extension might be to combine frequent trials in
simulation with infrequent trials in hardware~\citep{bongard}.  The simulation would produce the
necessary volume of trials to allow the learning methods to be
effective, and the hardware trials would serve to continuously ground
and refine the simulator.  One could also guide evolution to the most
fertile territory by penalizing gaits that produced large
discrepancies between simulation and reality~\citep{koos2010crossing}. Another natural extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands. All of these approaches would likely improve the quality of automatically generated gaits for legged robots, which will hasten the day that humanity can benefit from their vast potential.

>>>>

Because all trials were done in hardware, it was difficult to gather
the many trials that would be necessary to properly rank the methods statistically.
One direction for future work could be to obtain many more trials. However, a more effective extension might be to combine frequent trials in
simulation with infrequent trials in hardware~\citep{bongard}.  The simulation would produce the
necessary volume of trials to allow the learning methods to be
effective, and the hardware trials would serve to continuously ground
and refine the simulator.  One could also guide evolution to the most
fertile territory by penalizing gaits that produced large
discrepancies between simulation and reality~\citep{koos2010crossing}. Another extension would be to allow gaits that sensed the position of the robot and other variables to enable the robot to adjust to its physical state, instead of providing an open-loop sequence of motor commands. All of these approaches would likely improve the quality of automatically generated gaits for legged robots, which will hasten the day that humanity can benefit from their vast potential.

### 2011-04-14 paper_ecal/hyperNeatMethod.tex
Cleaned a few things up
More edits from reviews, made figures larger
more forging
more more forging

\subsection{HyperNEAT Gait Generation \& Learning}
\seclabel{hyperNeatMethod}

@@

\acmFig{hyperneatExplanation.png}{.8}{HyperNEAT produces ANNs from
  CPPNs. ANN weights are specified as a function of the geometric
  coordinates of the source node and the target node for each
  connection. The coordinates of these nodes and a constant bias are
  iteratively passed to the CPPN to determine each connection
  weight. The CPPN has two output values, which specify the weights
  for each connection layer as shown.  Figure from
  \cite{clune2011performance}.}

HyperNEAT is capable of exploiting the geometry of a problem because
the link values between nodes in the final ANN phenotype are a
function of the geometric positions of those nodes~\citep{stanley2009hypercube, clune2009sensitivity,
  clune2011performance}. In the case of quadruped locomotion, this
property has previously been shown to help HyperNEAT produce gaits with front-back, left-right,
and four-way symmetries~\citep{clune2009evolving,
  clune2011performance}.

@@

The ANN configuration follows previous studies that evolved quadruped
gaits with HyperNEAT in simulation~\citep{clune2011performance,
  clune2009evolving}, but was adapted to accommodate the physical robot used in this paper. Specifically, the ANN has a fixed topology (i.e. the number of nodes does not evolve) that consists of three $3 \times 4$
Cartesian grids of nodes forming input, hidden, and output
layers (\figref{SpiderANN.jpg}). Adjacent layers were allowed to be completely connected, meaning that there
could be $(3 \times 4)^2= 288$ links in each ANN (although evolution can set weights to 0, functionally eliminating the connection). The inputs to the
substrate were the current \emph{commanded} angles of each of the 9 joints of the robot and a sine and cosine wave (to facilitate the
production of periodic behaviors).  The sine and cosine waves had a period of about half a second.

@@

\acmFig{SpiderANN.jpg}{.7}{ANN configuration for HyperNEAT Runs. The
  first two columns of each row of the input layer receive information
  about a single leg (the current angle of its two joints). The final
  column provides a sine and cosine wave to enable periodic movements
  and the angle of the center joint. Evolution determines the function
  of the hidden-layer nodes. The nodes in the output layer specify new
  joint angles for each respective joint. The unlabeled nodes in the
  input and output layers are ignored. Figure adapted from
  \cite{clune2011performance}.}\label{hyperneatLayout}

>>>>

\subsection{HyperNEAT Gait Generation and Learning}
\seclabel{hyperNeatMethod}

@@

\acmFig{hyperneatExplanation.png}{1}{HyperNEAT produces ANNs from
  CPPNs. ANN weights are specified as a function of the geometric
  coordinates of the source node and the target node for each
  connection. The coordinates of these nodes and a constant bias are
  iteratively passed to the CPPN to determine each connection
  weight. The CPPN has two output values, which specify the weights
  for each connection layer as shown.  Figure from
  \cite{clune2011performance}.}

HyperNEAT can exploit the geometry of a problem because
the link values between nodes in the ANN phenotype are a
function of the geometric positions of those nodes~\citep{stanley2009hypercube, clune2009sensitivity,
  clune2011performance}. For quadruped locomotion, this
property has been shown to help HyperNEAT produce gaits with front-back, left-right,
and four-way symmetries~\citep{clune2009evolving,
  clune2011performance}.

@@

The ANN configuration follows previous studies that evolved quadruped
gaits with HyperNEAT in simulation~\citep{clune2011performance,
  clune2009evolving}, but was adapted to accommodate the physical robot in this paper. Specifically, the ANN has a fixed topology (i.e. the number of nodes does not evolve) that consists of three $3 \times 4$
Cartesian grids of nodes forming input, hidden, and output
layers (\figref{SpiderANN.jpg}). Adjacent layers were allowed to be completely connected, meaning that there
could be $(3 \times 4)^2= 288$ links in each ANN (although evolution can set weights to 0, functionally eliminating the connection). The inputs to the
substrate were the angles \emph{requested} in the previous time step for each of the 9 joints of the robot (recall that gaits are open-loop, so actual joint angles are unknown) and a sine and cosine wave (to facilitate the
production of periodic behaviors).  The sine and cosine waves had a period of about half a second.

@@

\acmFig{SpiderANN.jpg}{.7}{ANN configuration for HyperNEAT Runs. The
  first two columns of each row of the input layer receive information
  about a single leg (the angles requested in the previous time step for its two joints). The final
  column provides the previously requested angle of the center joint and, to enable periodic movements, a sine and cosine wave. Evolution determines the function
  of the hidden-layer nodes. The nodes in the output layer specify new
  joint angles for each respective joint. The unlabeled nodes in the
  input and output layers are ignored. Figure adapted from
  \cite{clune2011performance}.}\label{hyperneatLayout}

### 2011-04-14 paper_ecal/results.tex
more more forging

\acmFig{std_error}{1}{Average results ($\pm $ SE) for each of the
  parameterized learning methods, computed over three separately
  initialized runs.  Linear regression found the fastest overall gait
  and had the highest average, followed by Nelder-Mead simplex. Other
  methods did not outperform a random strategy.}

@@

The results for the gaits evolved by HyperNEAT are shown in
\figref{hnResults} and \tabref{results}.  A total of 540 evaluations
were performed for HyperNEAT, 180 in each of three runs.
Overall the HyperNEAT gaits
were the fastest by far, beating all the parameterized models when
comparing either average or best gaits.  We believe that this is
because HyperNEAT was allowed to explore a much richer space of
motions, but did so while still utilizing symmetries when
advantageous.  The single best gait found during this study had a
speed of 45.72 body lengths / minute, 66\% better than the best
non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
HyperNEAT gait that had high fitness.  The pattern of motion is both
complex (containing multiple frequencies and repeating patterns across
time) and regular, in that patterns of multiple motors are
coordinated.

The evaluation of the gaits produced by HyperNEAT was more noisy than
for the parameterized gaits, which made learning difficult. For
example, we tested an example HyperNEAT generation-champion gait 11
times and found that its mean performance was 26 body lengths per
minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
effective HyperNEAT gaits were not preserved across generations
because if performance in one trial was poor, the genome was unlikely
to be selected for. The HyperNEAT learning curve would be smoother if
the noise in the evaluations could be reduced or more than one
evaluation per individual could be afforded.

\acmFig{hnResults}{1}{Average fitness ($\pm $ SE) of the highest
  performing individual in the population for each generation of
  HyperNEAT runs. The fitness of many high-performing HyperNEAT gaits
  were halved if the gait overly stressed the motors (see text),
  meaning that HyperNEAT's true performance without this penalty would
  be much higher.  Each generation involved nine evaluations, one for
  each of the nine individuals, so the total evaluations per HyperNEAT
  run was 180.}

@@

\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Example of one
  high-performance gait produced by HyperNEAT showing
  commands for each of nine motors.  Note the complexity of the motion
  pattern. Such patterns were not possible with the parameterized
  SineModel5, nor would they likely be contrived by a human designing
  a different low-dimensional parameterized motion model.}

>>>>

\acmFig{std_error}{1}{Average results ($\pm $ SE) for the
  parameterized learning methods, computed over three separately
  initialized runs.  Linear regression found the fastest overall gait
  and had the highest average, followed by Nelder-Mead simplex. Other
  methods did not outperform a random strategy.}

@@

The results for the gaits evolved by HyperNEAT are shown in
\figref{hnResults} and \tabref{results}.  A total of 540 evaluations
were performed for HyperNEAT (180 in each of three runs).
Overall the HyperNEAT gaits
were the fastest by far, beating all the parameterized models when
comparing either average or best gaits.  We believe that this is
because HyperNEAT was allowed to explore a much richer space of
motions, but did so while still utilizing symmetries when
advantageous.  The single best gait found during this study had a
speed of 45.72 body lengths / minute, 66\% better than the best
non-HyperNEAT gait and 8.9 times faster than the hand-coded gait.
\figref{neat_110115_211410_00000_002_filt_zoom} shows a typical
HyperNEAT gait that had high fitness.  The pattern of motion is both
complex (containing multiple frequencies and repeating patterns across
time) and regular, in that patterns of multiple motors are
coordinated.

The evaluation of the gaits produced by HyperNEAT was more noisy than
for the parameterized gaits, which made learning difficult. For
example, we tested an example HyperNEAT generation-champion gait 11
times and found that its mean performance was 26 body lengths per
minute ($\pm$ 13 SD), but it had a max of 38 and a min of 3.  Many
effective HyperNEAT gaits were not preserved across generations
because a single poor-performing trial could prevent their selection. The HyperNEAT learning curve would be smoother if
the noise in the evaluations could be reduced or more than one
evaluation per individual could be afforded.

\acmFig{hnResults}{1}{Average fitness ($\pm $ SE) of the highest
  performing individual in the population for each generation of
  HyperNEAT runs. The fitness of many high-performing HyperNEAT gaits
  were halved if the gait overly stressed the motors (see text),
  meaning that HyperNEAT's true performance without this penalty would
  be even higher.}

@@

\acmFig{neat_110115_211410_00000_002_filt_zoom}{1}{Example of one
  high-performance gait produced by HyperNEAT showing
  commands for each of nine motors.  Note the complexity of the motion
  pattern. Such patterns were not possible with the parameterized
  SineModel5, nor would they likely result from a human designing
  a different low-dimensional parameterized motion model.}
